\section*{Foreword}

Robotics, an inherently multidisciplinary field, has witnessed unprecedented advancements since its inception in the 1960s. 
Over the decades, numerous disciplines have shown immense promise in tackling the challenges of creating autonomous systems. 
Yet, more than sixty years after the debut of Unimate, robots have still not fully integrated into the rich, unstructured, and dynamic world we humans inhabit.
This tutorial takes a clear stance in the debate on whether modern Machine 
Learning can play a pivotal role in the development of 
autonomous robot systems: we believe this to be the case.

However, we also hold that the wealth of research from both academia and industry in classical robotics over the past six decades is too valuable to be cast aside for purely learning-based methods. 
However, the interplay between classical robotics and modern machine learning is still in its nascent stages, with the path to integration yet to be clearly defined. 
While this tutorial focuses on learning-based approaches, we acknowledge the immense potential in combining these two techniques. 
Our goal in here is to present what we consider the most relevant and impactful approaches within robot learning today, and we warmly extend an invite to collaborate to expand the breadth of this work.

This tutorial covers a diverse range of topics in robotics and deep learning. Our aim is to provide a comprehensive conceptual guide, which has necessarily required us to omit certain details for the sake of clarity and focus.

This tutorial\dots
\begin{itemize}
    \item Does \emph{not} aim to be a comprehensive guide to robotics or manipulation:~\citet{sicilianoSpringerHandbookRobotics2016} and~\citet{tedrakeRoboticManipulationPerception} do this better than we ever could.
    \item Does \emph{not} aim to be an introduction to statistical or deep learning:~\citet{shalev-shwartzUnderstandingMachineLearning2014} and~\citet{prince2023understanding} cover these subjects better than we ever could.
    \item Does \emph{not} aim to be a deep dive into Reinforcement Learning, Diffusion Models, or Flow Matching: invaluable works such as~\citet{suttonReinforcementLearningIntroduction2018},~\citet{nakkiranStepbyStepDiffusionElementary2024}, and~\citet{lipmanFlowMatchingGuide2024} do this better than we ever could.
\end{itemize}

Instead, our goal is to illuminate how these disparate ideas have converged to form the exciting field of modern robot learning. We aim to provide intuition about why these techniques have been so effective together, driving the unprecedented progress we see today. In this spirit, we follow the adage: "a jack of all trades is a master of none, \emph{but oftentimes better than a master of one}."

We sincerely hope this tutorial serves as a valuable starting point for your journey into robot learning